---
title: "fgeo 1.2.0 release candidate"
output: rmarkdown::html_vignette
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.align = "center",
  out.width = "70%", 
  fig.width = 9.5, 
  fig.asp = 0.7
)

options(fgeo.quiet = TRUE)
```

[__fgeo__](https://forestgeo.github.io/fgeo/index.html) is a collection of R packages to analyze forest diversity and dynamics. We plan to announce it widely in early 2019. Before that, we would love you to [give us some feedback, and report issues](https://github.com/forestgeo/fgeo/issues).

This post highlights the main features. For more details please see the [Example](https://forestgeo.github.io/fgeo/index.html#example) and [Reference](https://forestgeo.github.io/fgeo/articles/siteonly/reference.html) sections.

### Setup

To install __fgeo__ see [Installation](https://forestgeo.github.io/fgeo/#installation). To use __fgeo__ run `library(fgeo)` in R.

```{r}
library(fgeo)
```

All __fgeo__ packages are now attached.

```{r}
search()
```

### Explore fgeo

Use [`fgeo_help()`](https://forestgeo.github.io/fgeo/reference/fgeo_help) to search documentation. It is similar to `utils::help.search()` but specific for __fgeo__.

### Access and manipulate data

__fgeo__ comes with small datasets for examples. When you use __fgeo__ you can also use many datasets directly.

```{r}
tree5
```

Other datasets are stored in R's library. You can access them with [`example_path()`](https://forestgeo.github.io/fgeo.x/reference/example_path).

```{r}
example_path("rdata")

dir(example_path("rdata"))
```

Larger datasets live online. You can download them with [`download_data()`](https://forestgeo.github.io/fgeo.x/reference/download_data).

```{r}
download_data("luquillo_tree5_random")
```

Use [`read_vft()`](https://forestgeo.github.io/fgeo.tool/reference/read_vft) to read your own ViewFullTable. Compared to `utils::read.csv()` it is easier to use and safer.

```{r}
(viewfulltable_file <- example_path("view/vft_4quad.csv"))

read_vft(viewfulltable_file)
```

### Plot

```{r}
# Using only two species to keep the example small
two_species <-  c("PREMON", "CASARB")

# The pipe operator (`%>%`) can make your code easier to read
census <- download_data("luquillo_stem6_random") %>% 
  filter(sp %in% two_species)
```

Use `autoplot()` to plot different classes of ForestGEO data. For example:

* [`autoplot()` species distribution](https://forestgeo.github.io/fgeo.plot/reference/autoplot.sp_elev)

```{r}
class(sp(census))

autoplot(sp(census))
```

* [`autoplot()` habitats](https://forestgeo.github.io/fgeo.plot/reference/autoplot.fgeo_habitat)

```{r}
# Using `data()` just for clarity
data("habitat")

class(habitat)

autoplot(habitat)
```

### Analyze

The available analyses include [abundance](https://forestgeo.github.io/fgeo.analyze/reference/abundance), [mortality](https://forestgeo.github.io/fgeo.analyze/reference/recruitment_ctfs), and [species-habitat association](https://forestgeo.github.io/fgeo.analyze/reference/tt_test).

```{r}
census %>% 
  pick_main_stem() %>% 
  group_by(sp) %>% 
  abundance()
```

```{r}
data("tree5")
data("tree6")

as_tibble(
  mortality_ctfs(tree5, tree6)
)
```

```{r}
tree <- download_data("luquillo_tree5_random")
data("habitat")

summary(
  tt_test(tree, habitat)
)
```

[Get started with __fgeo__](https://forestgeo.github.io/fgeo/)
